---
title: "Outbreak dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
---

```{r setup, echo=FALSE}
pacman::p_load(rio, here, tidyverse, flexdashboard,       # load packages
               flextable, incidence2, epicontacts, DT, janitor)

linelist <- import(here("data", "case_linelists", "linelist_cleaned.rds")) # import data
```

## Column 1 {data-width=500}
### Summary and action items 

This report is for the Incident Command team of the fictional outbreak of Ebola cases. **As of `r format(max(linelist$date_hospitalisation, na.rm=T), "%d %B")` there have been `r nrow(linelist)` cases reported as hospitalized.**  

* Several previously-unaffected areas to the West are now reporting cases  
* Internal reviews suggest that better communication is needed between district and provincial level data management teams  
* Safe and dignified burial teams are reporting difficulties  

### Review data
#### Cases by hospital
```{r}
linelist %>%
  count(hospital) %>%
  adorn_totals() %>%
  rename("Hospital" = hospital,
         "Cases" = n) %>% 
  knitr::kable()
```


## Column 2 {data-width=500}
### Epidemic curve by age

```{r}
age_outbreak <- incidence(linelist, date_onset, "week", groups = age_cat)
plot(age_outbreak, fill = age_cat, col_pal = muted, title = "") %>% 
  plotly::ggplotly()
```

### Transmission chain (select cases)
```{r}
# load package
pacman::p_load(epicontacts)

## generate contacts
contacts <- linelist %>%
  transmute(
    infector = infector,
    case_id = case_id,
    location = sample(c("Community", "Nosocomial"), n(), TRUE),
    duration = sample.int(10, n(), TRUE)
  ) %>%
  drop_na(infector)

## generate epicontacts object
epic <- make_epicontacts(
  linelist = linelist,
  contacts = contacts,
  id = "case_id",
  from = "infector",
  to = "case_id",
  directed = TRUE
)

## subset epicontacts object
sub <- epic %>%
  subset(
    node_attribute = list(date_onset = c(as.Date(c("2014-06-30", "2014-06-01"))))
  ) %>%
 thin("contacts")

# temporal plot
plot(
  sub,
  x_axis = "date_onset",
  node_color = "outcome",
  col_pal = c(Death = "firebrick", Recover = "green"),
  arrow_size = 0.5,
  node_size = 13,
  label = FALSE,
  height = 700,
  width = 700
)
```
